Controlling the dewetting morphologies of thin liquid films by switchable substrates

### 2022-11-25 Reply_RefereeB.tex

We thank the Referee for his/her thorough reading of our manuscript and detailed report.
We are happy to read that the referee acknowledges the amount of work we did and thinks it could be fitting for PRL after revision.
In what follows we provide detailed answers to all Referee's questions and we discuss all the changes made to the manuscript
(which are typeset in red colour).

\begin{itemize}

\item[ \textbf{\underline{Comment 1.}}]
{
The model disjoining pressure, Eq. 3 with a tail decaying as the
inverse squared power seems odd. I am more familiar with the expected
van der Waals tail with a decay as the inverse cube power. Is there
any reason for this choice?

\item[ \textbf{Answer}]
{
The Referee is right in pointing out that, indeed, the $h^{-3}$ decay 
for the attractive part of the disjoining pressure is more straightforward 
to understand as it can be derived directly from an intermolecular potential of Lennard-Jones type. Nevertheless, it is also true that other forms
can be encountered in the literature, depending on the actual interactions
\cite{SCHWARTZ1998173,Mitlin,Teletzke}. Here, though, our choice of the $(3,2)$ pair of exponents is motivated more by computational reasons rather than by the will of 
simulating a specific liquid with given physical properties.
With such choice, in fact, the characteristic dewetting time 
is shorter, thus making the overall computational cost more manageable.
In fact, we have seen in a previous study~\cite{PhysRevE.104.034801} that,
except for this shift of the time scales, the dewetting dynamics remains 
qualitatively the same. To confirm this observation, we have run 
a new simulation (for a single pair of pattern wavelength and speed) 
with the $(9,3)$ exponents and show the results in figure \ref{fig:rivulets_9_3}.
\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{9-3_3-2_comp_Gamma_1-5.pdf}
    \caption{Minkowski metric $q_2$ as a function of time for 
    $\Gamma = 1.5$ and $\lambda = 256 h_0$ from two simulations with different disjoining 
    pressure exponents, $(9,3)$ and $(3,2)$ (as in the paper) respectively.}
    \label{fig:rivulets_9_3}
\end{figure}
We have added the following footnote in the paper commenting on this aspect:\\
\\
\textcolor{red}{Expressions of 
the general form $\Pi(h) = \gamma(1-\cos \theta)(n-1)(m-1)/(n-m)((h_{\ast}/h)^n - (h_{\ast}/h)^m)/h_{\ast}$
are frequently found in the literature, with various combination of exponents $(n,m)$ ($n>m$), depending
on the type of interactions among fluid molecules and between fluid and substrate~\cite{SCHWARTZ1998173,Mitlin,Teletzke}. A common choice, for instance, is the pair $(9,3)$, which 
can be derived from molecular interactions of Lennard-Jones type. Here, the use of the $(3,2)$ pair is motivated mainly by the computational advantage of keeping the characteristic dewetting time 
shorter. We have run also a single case ($\lambda=256 h_0$ and $\Gamma=1.5$) with $(9,3)$ and observed that 
the dynamics remains essentially the same, but for a shift in the time scale (see also~\cite{PhysRevE.104.034801}).}
}

@@

\item[ \textbf{\underline{Comment 2.}}]
The second major concern refers to the choice of slip length. 
The authors study a fluid-substrate model that has a thin film equilibrium thickness of 
$0.07$ length units, but the slip length is one unit, hence about $10$ times larger than the equilibrium film thickness. 
This seems unexpected, in view of the small contact angle chosen for the system. 
Could the authors justify this choice?
}

\item[ \textbf{Answer}]
{
As the Referee correctly remarks, indeed small contact angles are typically associated to small slip lengths. From a phenomenological point of view, 
though, having $\delta \sim h_0$ is still commonly considered in the weak/intermediate slip regime, as proven, for instance, 
by the formation of satellite 
droplets following rivulet breakup~\cite{peschka2019signatures} or 
in the rim shape in dewetting by hole nucleation~\cite{fetzer2007quantifying, munch2005lubrication} (see also Fig.~\ref{fig:slip}).
\begin{figure}
    \centering
    \includegraphics [width=0.6\textwidth]{slip_measure.pdf}
    \caption{Dewetting rim profiles for various slip parameters $\delta$; notice the presence of a dimple behind the front, typical of 
    small to vanishing slip, which is preserved for slip lengths up to 
    $\delta \sim 10 h_0$.}
    \label{fig:slip}
\end{figure}
Our choice of this slip length value was dictated essentially by the need 
to keep numerical stability even in the strongly time dependent case 
(large pattern speed), yet remaining in a small slip regime. 
Nevertheless, motivated by the Referee's concern and in order to check that 
picture remains qualitatively the same, we have run a simulation with 
a slip length four times smaller. The results from the analysis of the Minkowski metric $q_2$, for two cases with low and high pattern speed, reveal that indeed
the droplet-to-rivulet transition is confirmed, with $\Gamma$ as the 
discriminating parameter (see figure \ref{fig:q2smallslip}).
\begin{figure}
    \centering
    \includegraphics [width=0.75\textwidth]{small_slip_two_Gammas.pdf}
    \caption{Minkowski metric $q_2$ as a function of time for two pattern speeds and slip velocity $\delta = 0.25 h_0$, signalling that the emergence of rivulets for $\Gamma>1$
    is preserved.}
    \label{fig:q2smallslip}
\end{figure}
We have commented on this when we give the value of the slip length, saying that:\\
\\
\textcolor{red}{To regularize the contact line divergence~\cite{huh1971hydrodynamic} we use a precursor layer thickness $h_{\ast}/h_0=0.07$ and a slip length $\delta/h_0 = 1$, 
a value which lies
within the weak/intermediate slip regime~\cite{peschka2019signatures,fetzer2007quantifying, munch2005lubrication} (we have tried also a value four times smaller, without observing 
any major qualitative change in the results).} 
}

@@

\item[ \textbf{\underline{Comment 3.}}]
{
The relevance of the manuscript relies on the significance of the substrate's switchable properties. 
Particularly, it is assumed that the wetting properties can be modulated on space but also on time. 
The authors have mentioned in the introduction that this can be potentially done. 
However, it would be desirable that the authors relate their model of switchable substrate with some specific experimental realization, where the actual length and time scales are mentioned.
}

\item[ \textbf{{Answer}}]
{
This is indeed a very interesting comment.
Although our approach, as theorists/numericists, can be seen more of {\it theoretical
engineering} and our picture would be, then, unlikely devoid of flaws, we are glad 
to try to envisage a potential experiment.
%We thank the Referee for bringing this up.
%First, we like to note that experiments are not our field of expertise and therefore %our envisioned realization could be flawed.

%The two relevant scales are $t_0$ and $q_0$ (or $2\pi/q_0$).
%Considering the experimental parameters from~\cite{becker2003complex,PhysRevLett.99.114503} %we get, 
%\begin{align*}
%    t_0 &\approx 300\,\text{s},\qquad~~~\qquad \frac{2\pi}{q_0}\approx 
%    400\,\text{nm}.
%\end{align*}
%The time scale is in the region of minutes and the wettability pattern wavelengths are %in the interval $(800\,\text{nm}, 4000\,\text{nm})$. 
A possibility would be to prepare a thin liquid film on a light responsive substrate~\cite{IchimuraEtAl_Science2000}, under the action of controlled 
external stimuli (a light emitter). 
An ideal candidate could be a digital multimirror device (DMD).
%The dimensions of the substrate should be roughly ten times the maximal pattern wavelength %($\approx 40\,\mu \text{m}$) to account for the periodicity.
%An ideal control of the external stimuli (light) would be a digital multimirror device (DMD).
This technology has effectively been used for thin film experiments and additive manufacturing~\cite{doi:10.1021/jp301092y, doi:10.1126/science.aax8760}, it allows for fast temporal modulations of the optical signal (with frequencies up to $\approx 16 \, \text{kHz}$)
with spatial resolution of $\approx 10 \times 10 \, \mu \text{m}^2$ (the size of a pixel).
%Using low intensity UV light $365\,\text{nm}$ and blue light %$436$nm~\cite{IchimuraEtAl_Science2000}, we assume that the radiation heat is negligible.
Considering as a reference, for instance, the system studied in ~\cite{becker2003complex,PhysRevLett.99.114503}, namely a $\sim 4 \, \text{nm}$ thick film 
of polystyrene deposited on an oxidized silicon wafer,  
we evaluate the retraction speed 
$U_{\Theta} = \frac{\Theta^3 \gamma}{9 \mu}$ to be 
$U_{\Theta} \approx 10^{-2} \, \mu \text{m}/\text{s}$. 
The (minimum) {\it pattern speed} can be 
estimated, from the pixel size with frame rate $\sim 1 s^{-1}$, as 
$v_{\theta} \sim 10 \, \mu \text{m}/\text{s}$, that would give $\Gamma \sim 10^3$, i.e. well within 
the rivulet regime ($\Gamma > 1$). We note, also, that both $U_{\Theta}$ and $v_{\theta}$ can 
be widely modulated: the former, playing either with the (temperature and molecular weight dependent) viscosity 
or tailoring the substrate to making it more hydrophobic (i.e. increasing the contact angle); the latter, tuning
the spatial resolution and frame rate of the DMD. We expect, thus, the range of achievable $\Gamma$'s to 
be feasibly extended both to very high ($\Gamma \gg 1$) and very low ($\Gamma \ll 1$) values. 
We have added the following footnote after the introduction of the contact angle pattern (Eq.(3) in the revised
manuscript) commenting on this aspect:\\
\\
\textcolor{red}{A laboratory experiment for the here mentioned approach consists of a light switchable substrate with a thin film cast on it~\cite{becker2003complex, IchimuraEtAl_Science2000}. 
The external stimulus, our moving pattern, can be realized by using a digital multimirror device (DMD)~\cite{doi:10.1021/jp301092y,doi:10.1126/science.aax8760}.} 
%\JH{Could one do this with a DMD as here https://doi.org/10.1021/jp301092y ? Then, one could use a transparent substrate and just do it from the bottom. I think a DMD + light responsive substrate could be a straightforward implementation of a time dependent wettability pattern.}

%\JH{We need to comment on the experimental realization in the paper. Can we summarize this idea to 1-2 sentences?}
}

@@

\item[ \textbf{\underline{Comment 4.}}]
{
An important parameter in the problem is the wave-vector $q_0$, which sets the correlation length in the problem. 
The authors should state what the actual value is relative to either the average film height, or the wave-length of modulations. 
Is there any reason to give the wave-length of modulations in units of the system size, i.e., are the results subject to finite size effects? Otherwise it appears more meaningful to describe the modulations in units of $q_0$ or the equilibrium film thickness.
}

\item[ \textbf{{Answer}}]
{
We apologize for not giving the value of $q_0 \approx 0.09$ (corresponding to an intrinsic correlation length 
$\lambda_s = 2\pi/q_0 \approx 70 h_0$), which is now provided in the revised manuscript. We also acknowledge that expressing the wavelength in terms of the size $L$
was not at all meaningful (no finite size effects are involved). In fact, all the lengths (including $L$) are expressed in units of the mean film thickness $h_0$ (and it was so also in the old version of the paper, as commented in the text below eq. (4)). In order to avoid any confusion, we explicitly write 
the values of all lengths in terms of $h_0$ in the revised version.
}

@@

\item[ \textbf{\underline{Comment 5.}}]
{
The thin film equation relies on a local approximation for the interface potential and disjoining pressure. 
Notice that to first order in non-local effects, the surface tension entering the thin film equation adopts a film thickness dependence which could potentially affect the dynamics. 
Particularly, as far as the parallel correlation length $q_0$ is referred, the non-local effects provide an upper bound equal to the inverse bulk correlation length. 
This could be a significant effect in polymer thin films (c.f. Benet et al. jpc-c 118, 22079, 2014 and the related discussion on dynamical effects Pahlavan et al., jfm, 845, 642, 2018).
}

\item[ \textbf{{Answer}}]
{
We guess that the Referee mentions the fact that the Laplacian of the thickness field, $\nabla^2 h$,
is just a small gradient approximation of the actual local curvature, such that the Laplace contribution to the film pressure should indeed read $\gamma \nabla^2 h/\sqrt{1+|\nabla h|^2}$. This is of course correct, but 
the condition $|\nabla h| \ll 1$ holds true in our system, owing to the very same lubrication approximation,
or, equivalently, to the small contact angles involved; the spinodal wavelength for a homogeneous substrate with
contact angle $\theta_0$ is $\lambda_s = 2\pi/q_0(\theta_0) \approx 70 h_0$. Therefore, the typical gradients
of the thickness field can be estimated as $|\nabla h| \sim h_0/\lambda_s \approx 1/70 \ll 1$ (notice 
that this is, somehow, a worst case scenario since, as we discuss in the text, in the patterned substrates
$\lambda \stackrel{>}{\sim} \lambda_s$).
We have added a footnote, when we introduce the film pressure, saying that:\\
\\
\textcolor{red}{The Laplace pressure is the product of the surface tension 
and the local curvature, therefore its full expression reads $p_L = -\gamma \nabla^2 h/\sqrt{1+|\nabla h|^2}$; 
since we are in the lubrication regime, though, whereby $|\nabla h| \ll 1$, the denominator can be safely neglected 
and the pressure reduces to the form involving only the Laplacian of the thickness field~\cite{Benet2014,alizadeh_pahlavan_cueto-felgueroso_hosoi_mckinley_juanes_2018}}, \\
\\
citing the related papers as a reference.
}
\end{itemize}

### 2022-11-25 Reply_RefereeC.tex

\item[ \textbf{\underline{Comment 2.}}]
{ 
The core of the manuscript is a numerical study which includes a
number of model parameters and specific choices of their values, say
$\theta_0$, $ \delta \theta$, $\delta$, $h_0$, 
the square pattern itself, the
orientation of the velocity of the contact angle, etc. Naturally,
these choices need to be made, but drawing conclusions from just a few
sets is, in my opinion, premature.
For example, I imagine that the rivulets will not form if $\delta\theta$ were too small, or too small relative to $\theta_0$. 
The authors focused on the importance of the speed of the pattern and on its wavelength but they left most other parameters fixed to a single value, the reasons for their choice being unclear.
}

\item[ \textbf{{Answer}}]
{ 
The Referee is right. Indeed there are of course several parameters 
that may influence the physics of the problem. 
Most of them, though, determine the dewetting dynamics even on static heterogeneous surfaces and are not specific of switchable substrates, with time-dependent properties.
For instance, as correctly identified by the Referee, $\delta \theta$ is one of those 
parameters. Of course, if $\delta \theta \rightarrow 0$ we do not expect to observe rivulets, as much as we do not expect to see arrays of droplets in the static case. Indeed, it is a known fact that if $\delta \theta$ were too small, the pattern driving would become too weak and the spinodal instability would set in as the main dewetting mechanism (see, e.g., \cite{KonnurPRL2000}). 
In this contribution, our aim is to single out the factor determining the novel effect found (the droplet-rivulet duality), namely the pattern time-dependence (and, hence, the 'wave' speed).  
Nevertheless, motivated by the Referee's remark, we have run a test case simulation in 
the expected rivulet-forming regime ($\Gamma=15$, $\lambda = 256 h_0$), reducing 
the heterogeneity parameter to $\delta \theta = 5^{\circ}$; 
the corresponding Minkowski's metric $q_2$ is reported 
in Fig. \ref{fig:dif_contrast} (together with the case $\delta \theta =10^{\circ}$, used throughout the 
paper, for comparison): we see that the $q_2$ metric attains the value $q_2 \approx 1$, signalling 
that the emergence of rivulets is preserved.
\begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{gamma15_deltaTheta.pdf}
    \caption{Comparison of the Minkowski's structure metric $q_2$ for 
    $\delta\theta = 5^{\circ}, 10^{\circ}$, 
    $\Gamma = 15$ and $\lambda = 256 h_0$.}
    \label{fig:dif_contrast}
\end{figure}
The rivulets lifetime, though, is shorter than the one for $\delta \theta = 10^{\circ}$.
We acknowledge, then, that a systematic study of the quantitative dependence of 
lifetimes on the full parameter space would be important, but we also that such a study
would deseve a dedicated, extended, work that goes somehow beyond the goal of the present paper and of a typical PRL contribution.
We have added the new figure and the relative discussion on $\delta \theta$ in the
Supplementary Material.\\
We agree, also, that it is appropriate to add further details in the text, therefore we commented in the revised version on the choice of the checkerboard pattern saying that:\\
\\
\textcolor{red}{We decorate the substrate with a checkerboard pattern, a common choice that generalizes the broken homogeneity of the stripes to two 
directions~\cite{KarguptaLangmuir2003,Jalali2018,Nagayama2020,Das2020}.}

}

\item[ \textbf{\underline{Comment 3.}}]
{ 
Also unclear is how does the substrate patterning scale $2\pi/q_\theta$ compare with the scale of homogeneous dewetting $2\pi/q_0$.
I tried hard to understand whether the different cases of $\lambda$ discussed in the paper agree with $2\pi/q_0$ but I was unable to do so. 
}

\item[ \textbf{{Answer}}]
{
We apologize for having lacked thoroughness not providing the numerical 
value of $q_0$ (of which we reported the expression, though) to compare with the 
other characteristic scales. Plugging in the values of $\theta_0$,
$h_0$ and $h_{\ast}$ we get $q_0 \approx 0.09 h_0^{-1}$, whence 
$\lambda_0 = 2\pi/q_0 \approx 70 h_0$, which corresponds approximately to the shortest pattern wavelength considered. 
In fact, it is known that for a periodic chemical pattern to be effective in 
confining droplets over the more hydrophilic patches, the pattern wavelength should 
not be smaller than the spinodal length scale on the less wettable substrate
($\lambda_l \approx 50$ in our case) \cite{KarguptaJCP2002,KarguptaLangmuir2000,Nisato1999,Karim1998}.
We have added comments on this important point in the revised manuscript, when discussing the accessible ranges of pattern wavelength (see also the answer to the comment 5).
}

\item[ \textbf{\underline{Comment 4.}}]
{  
How would the film behave if the dynamics of the film were controlled primarily by spinodal dewetting and not by the contact-angle pattern?
At the same time, if the two characteristic lengthscales ($=$ that of the spinodal dewetting and that of the patterned substrate) are suitably synchronized, perhaps rivulets can be rendered permanent rather than transient? Or is the dewetting, once initiated, immanently irreversible even for a
time-dependent contact-angle pattern like the one studied in LM17723?
}

\item[ \textbf{{Answer}}]
{
The pattern-driven instability overwhelms the spinodal instability because it is active even on an initially flat film 
(i.e., where the initial perturbation $|\delta h_0| \rightarrow 0$) 
\cite{KonnurPRL2000,KarguptaLangmuir2000,KarguptaPRL2001}; ideally one would need to have a vanishing 
contact angle mismatch in order to see the spinodal dewetting, namely $\delta \theta \sim \delta h_0/h_0$,
which is also somehow obvious, but it would eventually deprive the pattern of meaning. 
The emergence of rivulets is observed even for $\lambda \approx \lambda_s$, as we show 
in Fig. \ref{fig:q2_difflambda}, but their nature is still metastable. 
\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{lam_256-128-64-51.pdf}
    \caption{Minkowski's structure metric $q_2$ for three different wavelengths $\lambda=64 h_0$, $\lambda=128 h_0$ and $\lambda=256 h_0$ (as in figure 4 of the main text) and for $\Gamma=1.5$ (all other parameters are as in the main text). The time interval during which $q_2 \approx 1$ signals the emergence of rivulets, also for 
    $\lambda = 64 h_0$ which is comparable with the spinodal wavelength $\lambda_s \approx 70 h_0$}
    \label{fig:q2_difflambda}
\end{figure}
We have included also this new figure in the Supplementary Material.
}

\item[ \textbf{\underline{Comment 5.}}]
{ 
Many of the numerical results are supported by scaling estimates. I
like these arguments, but the agreement is hardly convincing. For
example, the linear and the square scaling laws in Fig. 2 each cover a
small range (much less than a tenfold variation) and so it is hard to
see whether they really apply. 
The same goes for the rivulet lifetime in Fig. 5 where the trend is OK but most likely a scaling law different from $\log(v_{\theta})$ would do as well in the range of $\tau_{riv}$ covered by the numerical data.
}

\item[ \textbf{{Answer}}]
{ 
We are well aware that the parameter ranges are not too wide, but 
of course our method (as any other) has to cope with limitations. 
For example, the range of achievable wavelengths is limited from below 
by the need to stay above the spinodal characteristic length (as explained in the answer to comment 3), and from above by the size of the system (i.e. by computing resources). As for the rivulet lifetimes, analogously, for reasons intrinsic to the stability of the numerical method, much higher pattern speeds cannot be reached 
(in particular one has to comply with the condition $\Gamma \ll 5 \times 10^2$). We have added the following
words of caution in the revised manuscript:\\
\\
\textcolor{red}{A few comments on the ranges are noteworthy at this point. The wavelengths are obviously limited from above by the size of the computational domain. On the other extreme, too short wavelengths are prevented by
the energetic cost of accomodating droplets on too
small patches, since the local contact angle (that increases with decreasing $\lambda$) would possibly exceed the largest substrate contact angle, such that the patterning 
itself would loose meaning. 
It has been, in fact, proven, both theoretically and experimentally, that 
for the confinement to be effective, the pattern characteristic size should 
not be much smaller than the spinodal length scale on the less wettable substrate
\cite{KarguptaJCP2002,KarguptaLangmuir2000,Nisato1999,Karim1998}.
We keep, therefore, $\lambda \stackrel{>}{\sim} \lambda_s \approx 70 h_0$. 
The pattern speeds, on the other hand, are limited by the intrinsic LB bound of the lattice speed of sound
 \cite{succi}.}\\
\\
We tend to disagree, instead, with the comment that any other scaling 
law would be equivalently good. Essentially because the logarithm scaling is 
not a fit, it is the result of a derivation based on plausible theoretical arguments, that 
is 'OK' with the numerical data.
%We think, anyhow, that the Referees' concerns are legitimate and we have, therefore, added 
%comments of caution about the scaling laws in the revised text.
Let us remark, though, that even in real microfluidic systems, 
one cannot expect too wild (and wide) variation of lengths and velocities.
In fact, we would be glad if our pioneering study might motivate further work 
in this direction, hopefully confirming our findings on an extended parameter 
range. Of course, that is more likely to happen if our message is conveyed
by a contribution in a high impact journal such as PRL. 
}
